---
embed-resources: true
code-tools: true
echo: true
---

# Part 1: Setup

## GitHub Workflow

Set up your GitHub workflow (either using the method of creating a repository and using Version Control to set up your project or vice versa using the `usethis` package commands we have learned).

Use appropriate naming conventions for your project (see Code Style Guide), e.g. lab-3-ggplot2.

Your project folder should contain the following:

-   `.Rproj`\
-   `lab-3-student.qmd`\
-   `data` folder
    -   `surveys.csv`\
-   rendered document (`.md`)

You will submit a link to your GitHub repository with all content.

## Seeking Help

Part of learning to program is learning from a variety of resources. Thus, I expect you will use resources that you find on the internet. There is, however, an important balance between copying someone else's code and *using their code to learn*. Therefore, if you use external resources, I want to know about it.

-   If you used Google, you are expected to "inform" me of any resources you used by **pasting the link to the resource in a code comment next to where you used that resource**.

-   If you used ChatGPT, you are expected to "inform" me of the assistance you received by (1) indicating somewhere in the problem that you used ChatGPT (e.g., below the question prompt or as a code comment), and (2) downloading and attaching the `.txt` file containing your **entire** conversation with ChatGPT. ChatGPT can we used as a "search engine", but you should not copy and paste prompts from the lab or the code into your lab.

Additionally, you are permitted and encouraged to work with your peers as you complete lab assignments, but **you are expected to do your own work**. Copying from each other is cheating, and letting people copy from you is also cheating. Please don't do either of those things.

## Lab Instructions

The questions in this lab are noted with numbers and boldface. Each question will require you to produce code, whether it is one line or multiple lines.

This document is quite plain, meaning it does not have any special formatting. As part of your demonstration of creating professional looking Quarto documents, **I would encourage you to spice your documents up (e.g., declaring execution options, specifying how your figures should be output, formatting your code output, etc.).**

## Setup

In the code chunk below, load in the packages necessary for your analysis. You should only need the `tidyverse` package for this analysis.

```{r}
#| label: setup                                                 # <1>
#| echo: true
#| output: false  
#| code-annnotations: hover                                     # <2>

library(tidyverse)                                                                                                               
```
1. Source citation of link including specifications used to format cell options: <https://quarto.org/docs/reference/cells/cells-knitr.html>
2. Source citation of link for formatting the citing of within-code annotations: <https://quarto.org/docs/authoring/code-annotation.html>



# Part 2: Data Context

The Portal Project is a long-term ecological study being conducted near Portal, AZ. Since 1977, the site has been used to study the interactions among rodents, ants, and plants, as well as their respective responses to climate. To study the interactions among organisms, researchers experimentally manipulated access to 24 study plots. This study has produced over 100 scientific papers and is one of the longest running ecological studies in the U.S.

We will be investigating the animal species diversity and weights found within plots at the Portal study site. The data are stored as a comma separated value (CSV) file. Each row holds information for a single animal, and the columns represent:

| Column            | Description                        |
|-------------------|------------------------------------|
| `record_id`       | Unique ID for the observation      |
| `month`           | month of observation               |
| `day`             | day of observation                 |
| `year`            | year of observation                |
| `plot_id`         | ID of a particular plot            |
| `species_id`      | 2-letter code                      |
| `sex`             | sex of animal ("M", "F")           |
| `hindfoot_length` | length of the hindfoot in mm       |
| `weight`          | weight of the animal in grams      |
| `genus`           | genus of animal                    |
| `species`         | species of animal                  |
| `taxon`           | e.g. Rodent, Reptile, Bird, Rabbit |
| `plot_type`       | type of plot                       |

## Reading the Data into `R`

We are going to use the `read_csv()` function to load in the `surveys.csv` dataset (stored in the data folder). For simplicity, name the data `surveys`. We will learn more about this function next week.

```{r}
#| label: load-data

surveys <- read_csv("data/surveys.csv")
glimpse(surveys)
```

**1. What are the dimensions (# of rows and columns) of these data?**

Using the glimpse() function, the output display shows that the 'surveys' dataframe contains 30,463 rows and 15 columns. 

**2. What are the data types of the variables in this dataset?**

The glimse() function output also informs that the 'surveys' dataframe contains data types of double and character. There is also one explicit column "date" containing date (time series) data.

# Part 3: Exploratory Data Analysis with **`ggplot2`**

`ggplot()` graphics are built step by step by adding new elements. Adding layers in this fashion allows for extensive flexibility and customization of plots.

To build a `ggplot()`, we will use the following basic template that can be used for different types of plots:

```{r}
#| eval: false
#| label: ggplot-template-code

ggplot(data = <DATA>,
       mapping = aes(<VARIABLE MAPPINGS>)) +
  <GEOM_FUNCTION>()
```

Let's get started!

## Scatterplot

**3. First, let's create a scatterplot of the relationship between `weight` (on the** $x$-axis) and `hindfoot_length` (on the $y$-axis).

```{r}
#| label: scatterplot
ggplot(data = surveys, 
       mapping = aes(
         x = weight, 
         y = hindfoot_length)
       ) +
  geom_point()
  
```

We can see there are **a lot** of points plotted on top of each other. Let's try and modify this plot to extract more information from it.

**4. Let's add transparency (`alpha`) to the points, to make the points more transparent and (possibly) easier to see.**

```{r}
#| label: Scatterplot pt-2 point-transparency

ggplot(data = surveys, 
       mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(mapping = aes(alpha = 0.5))

# Differentiate species by point color
ggplot(data = surveys, 
       mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(mapping = aes(alpha = 0.5, color = species))
```

Despite our best efforts there is still a substantial amount of overplotting occurring in our scatterplot. Let's try splitting the dataset into smaller subsets and see if that allows for us to see the trends a bit better.

**6. Facet your scatterplot by `species`.**

```{r}
#| label: Scatterplot pt-3 Faceting 

ggplot(data = surveys, 
       mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(mapping = aes(alpha = 0.5)) +
  facet_wrap(~species)
```

**7. No plot is complete without axis labels and a title. Include reader friendly labels and a title to your plot.**

```{r}
#| label: Scatterplot pt-4 Axis labels
#| echo: true
#| code-annnotations: hover

scttr_plot <- ggplot(data = surveys, 
       mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(mapping = aes(alpha = 0.5)) + 
  facet_wrap(~species) +
  labs(title = "Rodent Backfoot Length to Weight Ratios", 
                    x = "Weight (g)",
                    y = "Hindfoot Length (mm)",
       theme(plot.title = element_text(hjust = 0.5))) # Center title

scttr_plot + theme(plot.title = element_text(hjust = 0.5))      # <3>
 
```
3. Citation to reference graph label position adjustments: <https://ggplot2-book.org/themes#modifying-theme-components>


It takes a larger cognitive load to read text that is rotated. It is common practice in many journals and media outlets to move the $y$-axis label to the top of the graph under the title.

**8. Specify your** $y$-axis label to be empty and move the $y$-axis label into the subtitle.

```{r}
#| label: Scatterplot pt 5 axis label adjustment
#| echo: true

scttr_plot + labs(y = NULL, subtitle = "Hindfoot Length (mm)")
```


## Boxplots

**9. Create side-by-side boxplots to visualize the distribution of weight within each species.**
```{r}
#| label: Boxplot pt-1 weight by species
#| echo: true

ggplot(data = surveys, 
       mapping = aes(
         y = weight,
         x = species)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(size = 4))                   # <4>

```
4. This source was used to figure out how to adjust axis text size (as the defult size was too large and species labels were overlapping): <https://r-charts.com/ggplot2/axis/>


A fundamental complaint of boxplots is that they do not plot the raw data. However, with **ggplot** we can add the raw points on top of the boxplots!

**10. Add another layer to your previous plot that plots each observation using `geom_point()`.**

```{r}
#| label: Boxplot pt-2 overlay scatterplot
#| echo: true

ggplot(data = surveys, 
       mapping = aes(
         y = weight,
         x = species)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(size = 4)) +
  geom_point(aes(color = species))

ggplot(data = surveys) +
  geom_point(aes(x = weight, 
                 y = species, 
                 color = species)) +
  geom_boxplot(aes(x = weight, 
                   y = species))
```


Alright, this should look less than optimal. Your points should appear rather stacked on top of each other. To make them less stacked, we need to jitter them a bit, using `geom_jitter()`.

**11. Remove the previous layer and include a `geom_jitter()` layer instead.**

```{r}
#| label: Boxplot pt-3 jitter point overlay
#| echo: true

ggplot(data = surveys, 
       mapping = aes(
         y = weight,
         x = species)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(size = 4)) +
  geom_point(position = "jitter")                                # <5>
             
```
5. I used this source to format 'position' function and 'jitter' command appropriately: <https://r4ds.hadley.nz/layers.html>


That should look a bit better! But its really hard to see the points when everything is black.

**12. Set the `color` aesthetic in `geom_jitter()` to change the color of the points and add set the `alpha` aesthetic to add transparency.** You are welcome to use whatever color you wish! Some of my favorites are "springgreen4" and "steelblue4". Check them out on [R Charts](https://r-charts.com/colors/)

```{r}
#| label: Boxplot pt-4 add color & transparency to overlay 
#| echo: true

ggplot(data = surveys, 
       mapping = aes(
         y = weight,
         x = species)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(size = 3)) +
  geom_point(position = "jitter",
             aes(color = species, 
                   alpha = 0.5)) 
  
```


Great! Now that you can see the points, you should notice something odd: there are two colors of points still being plotted. Some of the observations are being plotted twice, once from `geom_boxplot()` as outliers and again from `geom_jitter()`!

**13. Inspect the help file for `geom_boxplot()` and see how you can remove the outliers from being plotted by `geom_boxplot()`. Make this change in your code!**

```{r}
#| label: Boxplot pt-5 remove boxplots outlier point overlays
#| echo: true

ggplot(data = surveys, 
       mapping = aes(
         y = weight,
         x = species)) +
  geom_boxplot(outliers = FALSE) +                               # <6>
  theme(axis.text.x = element_text(size = 3)) +
  geom_point(position = "jitter",
             aes(color = species, 
                   alpha = 0.5)) 
```
6. Code line hides boxplot's outlier points. Citation for code source inspired by: <https://ggplot2.tidyverse.org/reference/geom_boxplot.html>


Some small changes can make **big** differences to plots. One of these changes are better labels for a plot's axes and legend.

**14. Modify the** $x$-axis and $y$-axis labels to describe what is being plotted. Be sure to include any necessary units! You might also be getting overlap in the species names -- use `theme(axis.text.x = ____)` or `theme(axis.text.y = ____)` to turn the species axis labels 45 degrees.

```{r}
#| label: Boxplot pt-7 add labels
#| echo: true

ggplot(data = surveys, 
       mapping = aes(
         y = weight,
         x = species)) +
  geom_boxplot(outliers = FALSE) +                               
  theme(axis.text.x = element_text(size = 8)) +
  geom_point(position = "jitter",
             aes(color = species, 
                   alpha = 0.5)) +
  labs(x = "Species",
       y = "Weight (g)",
       ) +
  theme(axis.text.x = element_text(angle = 45),                  # <7>
        axis.text.y = element_text(angle = 45),                  # <7>
        legend.text = element_text(size = 10),                   # <7>
        legend.title = element_text(size = 10))                  # <7>
        
```
7. Adjust size of legend box by changing text size, citation reference: <https://www.spsanderson.com/steveondata/posts/2023-10-10/index.html>

Some people (and journals) prefer for boxplots to be stacked with a specific orientation! Let's practice changing the orientation of our boxplots.

**15. Now copy-paste your boxplot code you've been adding to above. Flip the orientation of your boxplots. If you created horizontally stacked boxplots, your boxplots should now be stacked vertically. If you had vertically stacked boxplots, you should now stack your boxplots horizontally!**

```{r}
#| label: rotated-boxplot
#| echo: true

ggplot(data = surveys, 
       mapping = aes(
         y = weight,
         x = species)) +
  geom_boxplot(outliers = FALSE) +  
  coord_flip() +                                                 # <8>
  theme(axis.text.x = element_text(size = 8)) +
  geom_point(position = "jitter",
             aes(color = species, 
                   alpha = 0.5)) +
  labs(x = "Species",
       y = "Weight (g)",
       ) +
  theme(axis.text.x = element_text(angle = 45),                  
        axis.text.y = element_text(angle = 45),                 
        legend.key.size = unit(0.1, "in")) # restage legend box size

```
8. Reorienting boxplot using coord_flip, because the orientation function was not working here. Citation reference: <https://ggplot2.tidyverse.org/reference/coord_flip.html>

Notice how vertically stacked boxplots make the species labels more readable than horizontally stacked boxplots (even when the axis labels are rotated). This is good practice!

# Lab 3 Submission

For Lab 3 you will submit the link to your GitHub repository. Your rendered file is required to have the following specifications in the YAML options (at the top of your document):

-   have the plots embedded (`embed-resources: true`)
-   include your source code (`code-tools: true`)
-   include all your code and output (`echo: true`)

**If any of the options are not included, your Lab 3 or Challenge 3 assignment will receive an "Incomplete" and you will be required to submit a revision.**

In addition, your document should not have any warnings or messages output in your HTML document. **If your HTML contains warnings or messages, you will receive an "Incomplete" for document formatting and you will be required to submit a revision.**

## References: 
1. *Quarto*. "Code Cells: Knitr". Accessed 25 Feb 2025. <https://quarto.org/docs/reference/cells/cells-knitr.html>.
2. *Quarto*. "Code Annotation". Accessed 25 Feb 2025. <https://quarto.org/docs/authoring/code-annotation.html>.
3. Hadley Wickham, Danielle Navarro, and Thomas Lin Pedersen. ggplot2: Elegant Graphics for Data Analysis (3rd Edition). "Modifying Theme Components". Designed using Quarto. Date Accessed 25 Feb 2025. <https://ggplot2-book.org/themes#modifying-theme-components>.
4. R Charts.  "ggplot2 axis titles, labels, ticks, limits, and scales". R Coder (author José Carlos Soage), 2024. Webpage accessed 25 Feb 2025. <https://r-charts.com/ggplot2/axis/>.
5. Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund. R for Data Science, 2nd edition. "Layers". Webpage accessed 25 Feb 2025. <https://r4ds.hadley.nz/layers.html>.
6. H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2016. "A box and whiskers plot (in the style of Tukey)". Developed by Posit. Site built with pkgdown 2.0.9.  <https://ggplot2.tidyverse.org/reference/geom_boxplot.html>.
7. Steven P. Sanderson II, MPH. 10 October, 2023. Steve's Data Tips and Tricks. "Resizing Legends in R". <https://www.spsanderson.com/steveondata/posts/2023-10-10/index.html>.
8. H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2016. "Cartesian coordinates with x and y flipped". Developed by Posit. Site built with pkgdown 2.0.9. <https://ggplot2.tidyverse.org/reference/coord_flip.html>.